On exponential sums arising from the classical doubling method

### 2025-04-08 main.tex
Moving section about gamma factors

\subsubsection{Gamma factors}\label{subsec:normalization-factor}
In order to make the computations more tolerable, it is beneficial to work with fully multiplicative functions. We introduce the doubling method gamma factors.

Let $\chi \colon \multiplicativegroup{\quadraticExtension} \to \multiplicativegroup{\cComplex}$ be a character.

We define a normalization factor as follows $$c_{\hermitianSpace}\left(\chi, \fieldCharacter\right) = \begin{dcases}
	\left(-\GaussSumSingleCharacter{\chi^{-2}}{\fieldCharacter}\right)^{\frac{\dim_{\finiteField} \hermitianSpace}{2}} & \quadraticExtension = \finiteField,\,\, \dim_{\finiteField} \hermitianSpace \text{ is even,}\\
	\chi\left(2\right) \left(-\GaussSumSingleCharacter{\chi^{-2}}{\fieldCharacter}\right)^{\frac{\dim_{\finiteField} \hermitianSpace - 1}{2}} & \quadraticExtension = \finiteField,\,\, \dim_{\finiteField} \hermitianSpace \text{ is odd,}	\\
	\left(-\GaussSumSingleCharacter{\chi^{-1}\restriction_{\multiplicativegroup{\finiteField}}}{\fieldCharacter}\right)^{\dim_{\quadraticExtension} \hermitianSpace} & \quadraticExtension \ne \finiteField.
\end{dcases}$$
Notice that if $\quadraticExtension \ne \finiteField$ then by the Hasse--Davenport relation $\GaussSumSingleCharacter{\chi^{-1}\restriction_{\multiplicativegroup{\finiteField}}}{\fieldCharacter}^{2} = \GaussSumSingleCharacter{\minusInvolutionMinusOne{\chi}} {\fieldCharacter_{\quadraticExtension}}$.

For an irreducible representation  $\tau$ of $\GL_k\left(\quadraticExtension\right)$, define $$\dblGammaFactor{\tau}{\chi}{\fieldCharacter} = \GaussSumScalar{\Contragradient{\tau} \times \involution{\chi}}{\fieldCharacter} \GaussSumScalar{\tau \times \chi}{\fieldCharacter}.$$
For an irreducible representation $\pi$ of $G$, define the \emph{doubling method gamma factor} $$\dblGammaFactorSpace{\hermitianSpace}{\pi}{\chi}{\fieldCharacter} = \frac{\posDblJacobiSumScalar{\pi}{\chi}}{c_{\hermitianSpace}\left(\chi, \fieldCharacter\right)}.$$

\subsection{Pairing with Deligne--Lusztig characters}
\calvin{old title: Computation of character sums attached to Deligne--Lusztig data}

@@

To summarize, we have
\begin{align*}
	& \sum_{\substack{x \in \NormOneGroup{2m}\\
			x \ne -1}} \theta \left(x\right) \chi\left(\aFieldNorm_{\finiteFieldExtension{2m} \slash \finiteField}\left(1 + x\right)\right) \\
		=& q^{\frac{m}{2}} \GaussSumSingleCharacter{\chi}{\fieldCharacter}^{2m} \GaussSumSingleCharacter{\chi^{-1}}{\fieldCharacter}^{2m} \GaussSumSingleCharacter{\chi^{-2}}{\fieldCharacter}^m \GaussSumCharacter{\left(\theta'\right)^{-1}}{\chi}{\fieldCharacter_{2m}},
\end{align*}
where $\theta'\left(t\right) = \theta\left(t^{1-q^m}\right)$.
Since $\tau\left(\chi^{-1}, \fieldCharacter\right) = \chi\left(-1\right) \conjugate{\tau\left(\chi, \fieldCharacter\right)}$ and since $\chi \ne 1$, we have that $\tau\left(\chi, \fieldCharacter\right) \tau\left(\chi^{-1}, \fieldCharacter\right) = \chi\left(-1\right)$. Combining this with the fact that $\left(\theta'\right)^{-1} = \left(\theta'\right)^{q^m}$ lies in the same Frobenius orbit as $\theta'$, we get the following theorem.
\begin{theorem}
	We have the identity $$\sum_{\substack{x \in \NormOneGroup{2m}\\
			x \ne -1}} \theta \left(x\right) \chi\left(\aFieldNorm_{\finiteFieldExtension{2m} \slash \finiteField}\left(1 + x\right)\right) = q^{\frac{m}{2}} \GaussSumSingleCharacter{\chi^{-2}}{\fieldCharacter}^m \GaussSumCharacter{\theta'}{\chi}{\fieldCharacter_{2m}}.$$
\end{theorem}

@@

Thus, we obtained the following.
\begin{theorem}
	We have the following identity $$\sum_{\substack{x \in \NormOneGroup{2m}\\
			x \ne -1}} \theta \left(x\right) \chi\left(\aFieldNorm_{\finiteFieldExtension{2m} \slash \finiteFieldExtension{2}}\left(1 + x\right)\right) = q^{\frac{m}{2}} \GaussSumSingleCharacter{\chi^{-1} \restriction_{\multiplicativegroup{\finiteField}}}{\fieldCharacter}^m \GaussSumCharacter{\theta'}{\chi}{\fieldCharacter_{2m}}.$$
\end{theorem}

@@

	Thus if $\pi$ is a subrepresentation of a parabolically induced representation as above, then $(T_H,\theta_H)$ is geometrically conjugate to $(T_{\GL_k\left(\quadraticExtension\right)} \times T_{H'}, \theta_{\GL_k\left(\quadraticExtension\right)} \times \theta_{H'})$ for $T_{\GL_k\left(\quadraticExtension\right)} \times H'$ a torus of $L = \GL_k\left(\quadraticExtension\right) \times H'$. As proved in \Cref{lem:indep_geo_conj}, $$g_{T \cap G}\left(\chi,\theta\restriction_{T \cap G},\psi\right) = g_{\left(T_{\GL_k\left(\quadraticExtension\right)} \times T_{H}'\right) \cap G}\left(\chi,\left(\theta_{\GL_k\left(\quadraticExtension\right)}\times \theta_{H'}\right)\restriction_{\left(T_{\GL_k\left(\quadraticExtension\right)} \times T_{H'}\right) \cap G},\psi\right).$$ The right hand of the last equality is multiplicative since Gauss sums are multiplicative by \eqref{eq:Gauss_multiplicitive}.
\end{proof}
\begin{proof}[Proof of \eqref{item:follows-for-cuspidal}]
By \Cref{thm:computation-of-doubling-gauss-sum-scalar-for-deligne-lusztig-characters} and \Cref{lem:indep_geo_conj}, we have that if $\RTThetaVirtual{T_H}{H}{\theta_H} = \sum_{\pi} c_{\pi} \pi$, where $\pi$ goes over all (equivalence classes of) irreducible representations of $H$ and $c_{\pi} \in \zIntegers$, then
$$\sum_{\pi} c_{\pi} \dblGammaFactorSpace{\hermitianSpace}{\pi}{\chi}{\fieldCharacter} = \GaussSumTorusCharacter{T'}{\theta'}{\chi}{\fieldCharacter} \RTGTheta{T_H}{H}{\theta_H}\left(1\right) = \GaussSumTorusCharacter{T'}{\theta'}{\chi}{\fieldCharacter} \left(\sum_{\pi} c_{\pi}\right).$$
Subtracting the equality for non-cuspidal $\pi$ established by (\ref{item:follows-for-non-cuspidal}), we are left with the equality
$$ \sum_{\pi \textnormal{ cuspidal}} c_{\pi} \dblGammaFactorSpace{\hermitianSpace}{\pi}{\chi}{\fieldCharacter} = \left(\sum_{\pi \textnormal{ cuspidal}} c_{\pi}\right)\GaussSumTorusCharacter{T'}{\theta'}{\chi}{\fieldCharacter}.$$
But by \Cref{prop:at-most-one-cuspidal-lusztig-series} there is at most one cuspidal $\pi$ appearing in our sum. Thus the proof of \Cref{thm:doubling-method-gamma-factor-for-deligne-lusztig} is complete.
\end{proof}

>>>>

\subsection{Pairing with Deligne--Lusztig characters}

@@

To summarize, we have
\begin{align*}
	& \sum_{\substack{x \in \NormOneGroup{2m}\\
			x \ne -1}} \theta \left(x\right) \chi\left(\aFieldNorm_{\finiteFieldExtension{2m} \slash \finiteField}\left(1 + x\right)\right) \\
		=& q^{\frac{m}{2}} \GaussSumSingleCharacter{\chi}{\fieldCharacter}^{2m} \GaussSumSingleCharacter{\chi^{-1}}{\fieldCharacter}^{2m} \GaussSumSingleCharacter{\chi^{-2}}{\fieldCharacter}^m \GaussSumCharacter{\left(\theta'\right)^{-1}}{\chi}{\fieldCharacter_{2m}},
\end{align*}
where $\theta'\left(t\right) = \theta\left(t^{1-q^m}\right)$.
Since $\tau\left(\chi^{-1}, \fieldCharacter\right) = \chi\left(-1\right) \conjugate{\tau\left(\chi, \fieldCharacter\right)}$ and since $\chi \ne 1$, we have that $\tau\left(\chi, \fieldCharacter\right) \tau\left(\chi^{-1}, \fieldCharacter\right) = \chi\left(-1\right)$. Combining this with the fact that $\left(\theta'\right)^{-1} = \left(\theta'\right)^{q^m}$ lies in the same Frobenius orbit as $\theta'$, we get the following theorem.
\begin{theorem}\label{thm:computation-norm-one-case}
	We have the identity $$\sum_{\substack{x \in \NormOneGroup{2m}\\
			x \ne -1}} \theta \left(x\right) \chi\left(\aFieldNorm_{\finiteFieldExtension{2m} \slash \finiteField}\left(1 + x\right)\right) = q^{\frac{m}{2}} \GaussSumSingleCharacter{\chi^{-2}}{\fieldCharacter}^m \GaussSumCharacter{\theta'}{\chi}{\fieldCharacter_{2m}}.$$
\end{theorem}

@@

Thus, we obtained the following.
\begin{theorem}\label{thm:computation-unitary-norm-one-case}
	We have the following identity $$\sum_{\substack{x \in \NormOneGroup{2m}\\
			x \ne -1}} \theta \left(x\right) \chi\left(\aFieldNorm_{\finiteFieldExtension{2m} \slash \finiteFieldExtension{2}}\left(1 + x\right)\right) = q^{\frac{m}{2}} \GaussSumSingleCharacter{\chi^{-1} \restriction_{\multiplicativegroup{\finiteField}}}{\fieldCharacter}^m \GaussSumCharacter{\theta'}{\chi}{\fieldCharacter_{2m}}.$$
\end{theorem}

\subsubsection{Gamma factors}\label{subsec:normalization-factor}
The multiplicativity property (\Cref{thm:multiplicativity-in-terms-of-gauss-sums}) and the computations performed above (Theorems \ref{thm:split-case-deligne-lusztig-computation}-\ref{thm:computation-unitary-norm-one-case}) suggest to normalize our Jacobi sums correctly so that they will satisfy a full multiplicativity property.

Let $\chi \colon \multiplicativegroup{\quadraticExtension} \to \multiplicativegroup{\cComplex} $ be a character. We define a normalization factor as follows $$c_{\hermitianSpace}\left(\chi, \fieldCharacter\right) = \begin{dcases}
	\left(-\GaussSumSingleCharacter{\chi^{-2}}{\fieldCharacter}\right)^{\frac{\dim_{\finiteField} \hermitianSpace}{2}} & \quadraticExtension = \finiteField,\,\, \dim_{\finiteField} \hermitianSpace \text{ is even,}\\
	\chi\left(2\right) \left(-\GaussSumSingleCharacter{\chi^{-2}}{\fieldCharacter}\right)^{\frac{\dim_{\finiteField} \hermitianSpace - 1}{2}} & \quadraticExtension = \finiteField,\,\, \dim_{\finiteField} \hermitianSpace \text{ is odd,}	\\
	\left(-\GaussSumSingleCharacter{\chi^{-1}\restriction_{\multiplicativegroup{\finiteField}}}{\fieldCharacter}\right)^{\dim_{\quadraticExtension} \hermitianSpace} & \quadraticExtension \ne \finiteField.
\end{dcases}$$
Notice that if $\quadraticExtension \ne \finiteField$ then by the Hasse--Davenport relation $\GaussSumSingleCharacter{\chi^{-1}\restriction_{\multiplicativegroup{\finiteField}}}{\fieldCharacter}^{2} = \GaussSumSingleCharacter{\minusInvolutionMinusOne{\chi}} {\fieldCharacter_{\quadraticExtension}}$. The factor on the left hand side appears in \Cref{thm:computation-unitary-norm-one-case}, while the one on the right hand side appears in \Cref{thm:multiplicativity-in-terms-of-gauss-sums}.

For an irreducible representation $\tau$ of $\GL_k\left(\quadraticExtension\right)$ define $$\dblGammaFactor{\tau}{\chi}{\fieldCharacter} = \GaussSumScalar{\Contragradient{\tau} \times \involution{\chi}}{\fieldCharacter} \GaussSumScalar{\tau \times \chi}{\fieldCharacter}.$$
For an irreducible representation $\pi$ of $H$, define the \emph{doubling method gamma factor} $$\dblGammaFactorSpace{\hermitianSpace}{\pi}{\chi}{\fieldCharacter} = \frac{\posDblJacobiSumScalar{\pi}{\chi}}{c_{\hermitianSpace}\left(\chi, \fieldCharacter\right)}.$$

@@

	Thus if $\pi$ is a subrepresentation of a parabolically induced representation as above, then $(T_H,\theta_H)$ is geometrically conjugate to $(T_{\GL_k\left(\quadraticExtension\right)} \times T_{H'}, \theta_{\GL_k\left(\quadraticExtension\right)} \times \theta_{H'})$ for $T_{\GL_k\left(\quadraticExtension\right)} \times H'$ a torus of $L = \GL_k\left(\quadraticExtension\right) \times H'$. As proved in \Cref{lem:indep_geo_conj}, $$g_{T \cap G}\left(\chi,\theta\restriction_{T \cap G},\psi\right) = g_{\left(T_{\GL_k\left(\quadraticExtension\right)} \times T_{H}'\right) \cap G}\left(\chi,\left(\theta_{\GL_k\left(\quadraticExtension\right)}\times \theta_{H'}\right)\restriction_{\left(T_{\GL_k\left(\quadraticExtension\right)} \times T_{H'}\right) \cap G},\psi\right).$$ The right hand of the last equality is multiplicative since Gauss sums are multiplicative by \eqref{eq:Gauss_multiplicitive}.
\end{proof}
\begin{proof}[Proof of \eqref{item:follows-for-cuspidal}]
By \Cref{thm:computation-of-doubling-gauss-sum-scalar-for-deligne-lusztig-characters} and \Cref{lem:indep_geo_conj}, we have that if $\RTThetaVirtual{T_H}{H}{\theta_H} = \sum_{\pi} c_{\pi} \pi$, where $\pi$ goes over all (equivalence classes of) irreducible representations of $H$ and $c_{\pi} \in \zIntegers$, then
$$\sum_{\pi} c_{\pi} \dblGammaFactorSpace{\hermitianSpace}{\pi}{\chi}{\fieldCharacter} = \GaussSumTorusCharacter{T'}{\theta'}{\chi}{\fieldCharacter} \RTGTheta{T_H}{H}{\theta_H}\left(1\right) = \GaussSumTorusCharacter{T'}{\theta'}{\chi}{\fieldCharacter} \left(\sum_{\pi} c_{\pi}\right).$$
Subtracting the equality for non-cuspidal $\pi$ established by (\ref{item:follows-for-non-cuspidal}), we are left with the equality
$$\sum_{\pi \textnormal{ cuspidal}} c_{\pi} \dblGammaFactorSpace{\hermitianSpace}{\pi}{\chi}{\fieldCharacter} = \GaussSumTorusCharacter{T'}{\theta'}{\chi}{\fieldCharacter} \left(\sum_{\pi \textnormal{ cuspidal}} c_{\pi}\right).$$
But by \Cref{prop:at-most-one-cuspidal-lusztig-series} there is at most one cuspidal $\pi$ appearing in our sum. Thus the proof of \Cref{thm:doubling-method-gamma-factor-for-deligne-lusztig} is complete.
\end{proof}
